---
title: "Run the paratuberculosis model in a cattle herd"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model_Example}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Install/load the library

If you have not already done so you must install the 'paratb' library
and its dependencies. There is a 'Makefile' in the root of the package
that contains the instructions to build a install the package. Then
you may load the package:

```{r, warning = FALSE}
library(paratb)
```
## Initialise the model.

In the model used to study spread of paratuberculosis in Sweden we
would use a list of herds in the population and the number of animals
in each herd and age category at the beginning of the study
period. This default model has 10,000 fake farms with 200 animals in each
farm, to change the starting state replace `u0` with another similarly
structured 'data.frame'. Here we initialise the model with the
default fake herds and specify that we want to model over 10 years,
with reported results every 90 days:

```{r, warning = FALSE}
model <- paratb(tspan = seq(1, 10*365, 90))
```


## Seed infection in 0.2% of the herds

In the study model we seeded infection 2% of animals in 0.2% of
herds. We will do the same seeding in this model:

```{r, warning = FALSE}
for(i in seq_len(ncol(model@u0)*0.002)) {
model <- seed_herd(model,
                   i,
                   "cow",
                   ceiling(model@u0["S3",i] * 0.02))
model <- seed_herd(model,
                   i,
                   "heifer",
                   ceiling(model@u0["S2",i] * 0.02))
model <- seed_herd(model,
                   i,
                   "calf",
                   ceiling(model@u0["S1",i] * 0.02))
}
```
## Add animal-movement events

In the study model, we used approximately 15 million industry-recorded
animal movements to reflect the real Swedish population over 13
years. For this model we will add 1 million random events to send
some animals around between herds.

```{r, warning = FALSE, cache=TRUE}
model <- random_events(model, 1000000)
```
Now we have an initialised model object, ready to run a spread
simulation and look at the results. First run the model:

```{r, eval=FALSE, echo=TRUE}
result <- run(model)
```
```{r, eval=TRUE, echo=FALSE}
set.seed(99)
result <- run(model)
```

Then extract and plot the number of infected herds over time:
```{r, warning = FALSE, results = 'asis', fig.width= 8, fig.height = 6}
df <- counti(result, "true")

plot(df$time,
     df$count,
     type = "s",
     ylim = c(0, max(df$count)),
     ylab = "Count of positive herds",
     xlab = "Time in days")
```


## Add surveillance to the model

In the study model we added surveillance events to the model to study
various herd selection strategies and numbers of herds tested per year
with a bulk milk ELISA for paratuberculosis. Now run the same model
but inject bulk milk sampling events in 1000 herds per year and
monitor the outcome of that compared to the disease.

```{r, eval=TRUE, echo=TRUE}
years <- max(model@tspan)/365

herds_for_surveillance <- sample(seq_len(ncol(model@u0)),
                                 1000 * years,
                                 replace = TRUE)

times_for_surveillance <- sample(min(model@tspan):max(model@tspan),
                                 length(herds_for_surveillance),
                                 replace = TRUE)

model <- add_surveillance_event(model,
                                i = herds_for_surveillance,
                                t = times_for_surveillance,
                                n = 1L)
```
Run the model again but with the surveillance additional surveillance events:

```{r, eval=FALSE, echo=TRUE}
result <- run(model)
```
```{r, eval=TRUE, echo=FALSE}
set.seed(99)
result <- run(model)
```
Extract the data we need about the herd status and the detection
status and plot it. In the study model we summarised this result over
many trajectories to determine the likely efficacy of the
surveillance. Here we will just look at a single trajectory:

```{r, warning = FALSE, results = 'asis', fig.width= 8, fig.height = 6}
df <- counti(result, "true")
df_bulk_milk <- counti(result, "milk")

plot(df$time,
     df$count,
     type = "s",
     ylim = c(0, max(df$count)),
     ylab = "Count of positive/detected herds",
     xlab = "Time in days")

lines(df_bulk_milk$time,
     df_bulk_milk$count,
     type = "s",
     col = "#939393",
     lty = 2)
```
